---
title: "DHT Mice"
author: "Victoria Sugrue"
date: 'Last updated: 5 August 2024'
output:
  word_document:
    toc: no
    toc_depth: '7'
  html_document:
    toc: no
    toc_float: yes
    toc_depth: 7
  pdf_document:
    toc: no
    toc_depth: 5
    df_print: kable
    latex_engine: xelatex
subtitle: ' '
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data preparation

```{r}
library(dplyr)
library(ggplot2)
library(cowplot)
library(Manu)
library(ggpubr)

data <- read.csv("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Revision/MouseModels_June2024.csv")
data$Beta_corrected <- data$Beta-0.042
capsuleDHTdata2 <- data %>% filter(Experiment=="CapsuleDHT")
capsuleDHTdata <- capsuleDHTdata2 %>% filter(!(Treatment=="DHTLOST"))
prenatalDHTdata <- data %>% filter(Experiment=="PrenatalDHT")
ARKOdata <- data %>% filter(Experiment=="ARKO")
VGlutdata <- data %>% filter(Experiment=="Vglut")

capsuleDHTdata <- capsuleDHTdata %>% mutate(Treatment = ifelse(Treatment == "DHT", "C-DHT", Treatment))
capsuleDHTdata$Treatment <- factor(capsuleDHTdata$Treatment, levels = c("Control", "C-DHT"))
```

# Capsule DHT mice

```{r}
capsule_ear <- capsuleDHTdata %>% filter(Tissue=="Ear")
capsule_tail <- capsuleDHTdata %>% filter(Tissue=="Tail")
capsule_muscle <- capsuleDHTdata %>% filter(Tissue=="Muscle")
capsule_liver <- capsuleDHTdata %>% filter(Tissue=="Liver")
capsule_kidney <- capsuleDHTdata %>% filter(Tissue=="Kidney")
```

# Ear

## Plot
```{r}
DHTear <- ggplot(capsule_ear, aes(x=Treatment, y=CG.methylation, colour=Treatment)) +
  geom_boxplot(varwidth=T) + 
  geom_point(size=5, alpha=4/6) +
  labs(title="Ear", 
       x="Treatment",
       y="Methylation (%)") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme_light() +
  theme(legend.position = "none") + 
  theme(text = element_text(size = 20))
DHTear
```

## Statistics 

### T test
```{r}
t.test(capsule_ear$CG.methylation~capsule_ear$Treatment)

```

### Correlation between methylation and weight
```{r}
ear_DHT <- capsule_ear %>% filter(Treatment=="C-DHT")
cor.test(ear_DHT$CG.methylation, ear_DHT$Weight)

ear_Control <- capsule_ear %>% filter(Treatment=="Control")
cor.test(ear_Control$CG.methylation, ear_Control$Weight)
```


# Tail

## Plot
```{r}
DHTtail <- ggplot(capsule_tail, aes(x=Treatment, y=CG.methylation, colour=Treatment)) +
  geom_boxplot(varwidth=T) + 
  geom_point(size=5, alpha=4/6) +
  labs(title="Tail", 
       x="Treatment",
       y="Methylation (%)") +
  theme_light() +
    scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme(legend.position = "none") + 
  theme(text = element_text(size = 20)) 
DHTtail
```

## Statistics 

### T test
```{r}
t.test(capsule_tail$CG.methylation~capsule_tail$Treatment)

```



### Correlation between methylation and weight
```{r}
tail_DHT <- capsule_tail %>% filter(Treatment=="C-DHT")
cor.test(tail_DHT$CG.methylation, tail_DHT$Weight)

tail_Control <- capsule_tail %>% filter(Treatment=="Control")
cor.test(tail_Control$CG.methylation, tail_Control$Weight)
```


# Muscle

## Plot
```{r}
DHTmuscle <- ggplot(capsule_muscle, aes(x=Treatment, y=CG.methylation, colour=Treatment)) +
  geom_boxplot(varwidth=T) + 
  geom_point(size=5, alpha=4/6) +
  labs(title="Muscle", 
       x="Treatment",
       y="Methylation (%)") +
  theme_light() +
    scale_colour_manual(values = get_pal("Takahe")[c(1,2,4)]) +
  theme(legend.position = "none") + 
  theme(text = element_text(size = 20)) 
DHTmuscle
```

## Statistics 

### T test
```{r}
t.test(capsule_muscle$CG.methylation~capsule_muscle$Treatment)
```



### Correlation between methylation and weight
```{r}
muscle_DHT <- capsule_muscle %>% filter(Treatment=="C-DHT")
cor.test(muscle_DHT$CG.methylation, muscle_DHT$Weight)

muscle_Control <- capsule_muscle %>% filter(Treatment=="Control")
cor.test(muscle_Control$CG.methylation, muscle_Control$Weight)
```


# Kidney

## Plot
```{r}
DHTkidney <- ggplot(capsule_kidney, aes(x=Treatment, y=CG.methylation, colour=Treatment)) +
  geom_boxplot(varwidth=T) + 
  geom_point(size=5, alpha=4/6) +
  labs(title="Kidney", 
       x="Treatment",
       y="Methylation (%)") +
  theme_light() +
    scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme(legend.position = "none") + 
  theme(text = element_text(size = 20))
DHTkidney
```

## Statistics 

### T test
```{r}
t.test(capsule_kidney$CG.methylation~capsule_kidney$Treatment)

```


### Correlation between methylation and weight
```{r}
kidney_DHT <- capsule_kidney %>% filter(Treatment=="C-DHT")
cor.test(kidney_DHT$CG.methylation, kidney_DHT$Weight)

kidney_Control <- capsule_kidney %>% filter(Treatment=="Control")
cor.test(kidney_Control$CG.methylation, kidney_Control$Weight)
```



# Liver

## Plot
```{r}
DHTliver <- ggplot(capsule_liver, aes(x=Treatment, y=CG.methylation, colour=Treatment)) +
  geom_boxplot(varwidth=T) + 
  geom_point(size=5, alpha=4/6) +
  labs(title="Liver", 
       x="Treatment",
       y="Methylation (%)") +
  theme_light() +
    scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme(legend.position = "none") + 
  theme(text = element_text(size = 20)) 
DHTliver
```

## Statistics 

### T test
```{r}
t.test(capsule_liver$CG.methylation~capsule_liver$Treatment)

```


### Correlation between methylation and weight
```{r}
liver_DHT <- capsule_liver %>% filter(Treatment=="C-DHT")
cor.test(liver_DHT$CG.methylation, liver_DHT$Weight)

liver_Control <- capsule_liver %>% filter(Treatment=="Control")
cor.test(liver_Control$CG.methylation, liver_Control$Weight)
```

# Plot together

```{r}
capsuleDHTdata$Tissue <- factor(capsuleDHTdata$Tissue, levels = c("Muscle", "Tail", "Kidney", "Ear", "Liver"))

# Plot
DHT_combined <- ggplot(capsuleDHTdata, aes(x=Tissue, y=CG.methylation, colour=Treatment)) +
  geom_boxplot(varwidth=T) + 
  geom_point(size=5, alpha=4/6, position=position_dodge(width=0.75)) +
  labs(title="  ", 
       x="Tissue",
       y="Methylation (%)") +
  theme_light() +
  geom_signif(y_position = c(95, 95, 95, 95), xmin = c(0.75, 1.75, 2.75, 3.75), xmax = c(1.25, 2.25, 3.25, 4.25), tip_length = 0.02, annotation = c("***", "***", "**", "***"), colour="black", textsize = 5) +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
theme(legend.position = c(0.87, 0.20),
        legend.background = element_rect(fill = "white", color = "black")) +  theme(text = element_text(size = 10)) +
  guides(colour = guide_legend(title = "Treatment", nrow = 1)) +
  ylim(30,100) 

DHT_combined
```




# Apply it to the clock
# Data preparation
```{r}
mouse_array_data <- read.csv("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Data/Mouse_ArrayData_Full_TailMuscle.csv")

# Setting AndroAge with predicted time of puberty onset = age minus 6 weeks, adjust to 0 where negative
mouse_array_data$YPMP <- mouse_array_data$Age-0.115068
mouse_array_data$YPMP[mouse_array_data$Sex=="F"] <- 0
mouse_array_data$YPMP[mouse_array_data$YPMP < 0] <- 0

# Change sex labels
mouse_array_data$Sex[mouse_array_data$Sex=="F"] <- "Female"
mouse_array_data$Sex[mouse_array_data$Sex=="M"] <- "Male"

# Subset by tissue
mouse_array_muscle <- mouse_array_data %>% filter(Tissue=="Muscle")
mouse_array_tail <- mouse_array_data %>% filter(Tissue=="Tail")
```

# I also need to add to my new data: 
- a column with the corrected beta value in it headed "cg21524116" in my new data
- AgeYears (base this off the existing AgeDays column)
- YPMP values

```{r}
capsule_tail$cg21524116 <- capsule_tail$Beta_corrected
capsule_tail$AgeYears <- capsule_tail$AgeDays/365
capsule_tail$YPMP <- capsule_tail$AgeYears-0.115068
capsule_tail$YPMP[capsule_tail$Sex=="Female"] <- 0
capsule_tail$YPMP[capsule_tail$YPMP < 0] <- 0
```


##### Tail #####

# Fit androgen clock to mouse tail
```{r}
male_mouse_array_tail <- mouse_array_tail %>% filter(Sex=="Male")

tail_model <- lm(YPMP ~ cg21524116, data=male_mouse_array_tail) 
summary(tail_model)
```

# Now, to apply it to the new data.

```{r}
capsule_tail$Model_AndroAge_Tailclock <- predict(tail_model, newdata = capsule_tail)

capsule_tail_model <- as.data.frame(Metrics::mdae(actual=capsule_tail$YPMP, predicted=capsule_tail$Model_AndroAge_Tailclock))
capsule_tail_model
# 0.485
```

```{r}
capsule_tail$AndroAgeAccel <- capsule_tail$Model_AndroAge_Tailclock-capsule_tail$YPMP
```


```{r}
capsule_tail_DHT <- capsule_tail %>% filter(Treatment=="C-DHT")
capsule_tail_control <- capsule_tail %>% filter(Treatment=="Control")
```


```{r}
Metrics::mdae(actual=capsule_tail_DHT$YPMP, predicted=capsule_tail_DHT$Model_AndroAge_Tailclock)
# 1.244

mean(capsule_tail_DHT$Model_AndroAge_Tailclock)
# 0.991
```

```{r}
Metrics::mdae(actual=capsule_tail_control$YPMP, predicted=capsule_tail_control$Model_AndroAge_Tailclock)
# 0.400

mean(capsule_tail_control$Model_AndroAge_Tailclock)
# -0.437
```
# Mann whitney p value
```{r}
wilcox.test(capsule_tail$Model_AndroAge_Tailclock~capsule_tail$Treatment)
# p=0.0003108
```

# Plot

```{r}
d2tail <- capsule_tail %>%
  dplyr::group_by(Treatment) %>%
  dplyr::summarise(
    mean = mean(Model_AndroAge_Tailclock, na.rm = TRUE),
    sd = sd(Model_AndroAge_Tailclock, na.rm = TRUE)
  )
```

```{r}
AndroAgeTail <- ggplot(d2tail, aes(x=Treatment, y=mean)) + 
  geom_bar(stat = "identity", fill="lightgrey", colour="black") + 
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.5) + 
  geom_point(capsule_tail, mapping = aes(x=Treatment, y=Model_AndroAge_Tailclock, colour=Treatment), size=4, alpha=1/2) +
  ylab("AndroAge (Years)") + theme_classic() + geom_hline(yintercept=0) + 
  theme_light() +
  ggtitle("Chronic DHT Tail") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme(legend.position="none", plot.title = element_text(hjust = 0.5))+
  geom_signif(y_position = c(2.15), xmin = c(1), xmax = c(2), tip_length = 0.02, annotation = c("***"), colour="black", textsize = 5) +
  theme(text = element_text(size = 10)) +
  ylim(-1, 2.7)
AndroAgeTail
```

##### Muscle #####

```{r}
capsule_muscle$cg21524116 <- capsule_muscle$Beta_corrected
capsule_muscle$AgeYears <- capsule_muscle$AgeDays/365
capsule_muscle$YPMP <- capsule_muscle$AgeYears-0.115068
capsule_muscle$YPMP[capsule_muscle$Sex=="Female"] <- 0
capsule_muscle$YPMP[capsule_muscle$YPMP < 0] <- 0
```

```{r}
male_mouse_array_muscle <- mouse_array_muscle %>% filter(Sex=="Male")

muscle_model <- lm(YPMP ~ cg21524116, data=male_mouse_array_muscle) 
summary(muscle_model)
```

# Now, to apply it to the new data.

```{r}
capsule_muscle$Model_AndroAge_Muscleclock <- predict(muscle_model, newdata = capsule_muscle)

capsule_data_model_muscleclock <- as.data.frame(Metrics::mdae(actual=capsule_muscle$YPMP, predicted=capsule_muscle$Model_AndroAge_Muscleclock))
capsule_data_model_muscleclock
# 0.957
```

```{r}
capsule_muscle$AndroAgeAccel <- capsule_muscle$Model_AndroAge_Muscleclock-capsule_muscle$YPMP
```


```{r}
capsule_muscle_DHT <- capsule_muscle %>% filter(Treatment=="C-DHT")
capsule_muscle_control <- capsule_muscle %>% filter(Treatment=="Control")
```


```{r}
as.data.frame(Metrics::mdae(actual=capsule_muscle_DHT$YPMP, predicted=capsule_muscle_DHT$Model_AndroAge_Muscleclock))
# 1.552

mean(capsule_muscle_DHT$Model_AndroAge_Muscleclock)
# 1.591
```

```{r}
as.data.frame(Metrics::mdae(actual=capsule_muscle_control$YPMP, predicted=capsule_muscle_control$Model_AndroAge_Muscleclock))
# 0.083

mean(capsule_muscle_control$Model_AndroAge_Muscleclock)
# -0.204
```
# Mann whitney p value
```{r}
wilcox.test(capsule_muscle$Model_AndroAge_Muscleclock~capsule_muscle$Treatment)
# p=0.0003108
```


# Plot

```{r}
d2muscle <- capsule_muscle %>%
  dplyr::group_by(Treatment) %>%
  dplyr::summarise(
    mean = mean(Model_AndroAge_Muscleclock, na.rm = TRUE),
    sd = sd(Model_AndroAge_Muscleclock, na.rm = TRUE)
  )
```

```{r}
AndroAgeMuscle <- ggplot(d2muscle, aes(x=Treatment, y=mean)) + 
  geom_bar(stat = "identity", fill="lightgrey", colour="black") + 
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.5) + 
  geom_point(capsule_muscle, mapping = aes(x=Treatment, y=Model_AndroAge_Muscleclock, colour=Treatment), size=4, alpha=1/2) +
  ylab("AndroAge (Years)") + theme_classic() + geom_hline(yintercept=0) + 
  theme_light() +
  ggtitle("Chronic DHT Muscle") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme(legend.position="none", plot.title = element_text(hjust = 0.5)) +
  geom_signif(y_position = c(2.15), xmin = c(1), xmax = c(2), tip_length = 0.02, annotation = c("***"), colour="black", textsize = 5) +
  theme(text = element_text(size = 10)) +
  ylim(-1, 2.7)
AndroAgeMuscle
```

# ARKO samples

```{r}
ARKOdata$FullDetailsBrief <- factor(ARKOdata$FullDetailsBrief, 
                                     levels = c("MaleWT", "FemaleWT", "MaleKO", "MaleKO+T"))

ARKO_plot_full <- ggplot(ARKOdata, aes(x=FullDetailsBrief, y=CG.methylation, colour=FullDetailsBrief)) +
  geom_boxplot(varwidth=T) + 
  geom_jitter(size=5, width=0.01, alpha=4/5) +
  labs(title="", 
       x="Treatment",
       y="Methylation (%)") +
  theme_light() +
  theme(legend.position = "none") + 
  #theme(text = element_text(size = 20)) +
  scale_colour_manual(values = get_pal("Takahe")[c(2,1,4,3)]) +
  geom_signif(y_position = c(80), xmin = c(1), xmax = c(2), tip_length = c(0.03, 0.03), annotation = c(""), textsize = 8) +
  geom_signif(y_position = c(93.5), xmin = c(1), xmax = c(3), tip_length = c(0.03, 0.03), annotation = c(""), textsize = 8) +
  geom_signif(y_position = c(98), xmin = c(1), xmax = c(4), tip_length = c(0.03, 0.03), annotation = c(""), textsize = 8) +
  geom_signif(y_position = c(84.5), xmin = c(3), xmax = c(4), tip_length = c(0.03, 0.03), annotation = c(""), textsize = 8) +
  geom_signif(y_position = c(89), xmin = c(2), xmax = c(3), tip_length = c(0.03, 0.03), annotation = c(""), textsize = 8) +
  ylim(25,100) + 
  geom_text(aes(x = 1.5, y = 81, label = "***"), size = 6, colour="black") +
  geom_text(aes(x = 2.5, y = 90, label = "*"), size = 6, colour="black") +
  geom_text(aes(x = 3.5, y = 85.5, label = "*"), size = 6, colour="black") +
  geom_text(aes(x = 2, y = 94.5, label = "***"), size = 6, colour="black") +
  geom_text(aes(x = 3, y = 99, label = "***"), size = 6, colour="black") 

ARKO_plot_full

ggsave("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Revision/FinalFigures/FigureSX_ARKOMethylation.png", ARKO_plot_full, width=8,height=5, dpi=1000)
```

# maybe this should be plot by age??
```{r}
ARKO_plot_age <- ggplot(ARKOdata, aes(x=AgeDays, y=CG.methylation, colour=FullDetailsBrief)) +
  geom_point(size=5, alpha=4/5) +
   labs(title="", 
       x="Age (Days)",
       y="Methylation (%)",
       colour="Group") +
  theme_light() +
  geom_smooth(method=lm, se=F) +
  facet_wrap(~FullDetailsBrief) +
  scale_colour_manual(values = get_pal("Takahe")[c(2,1,4,3)]) +
  theme(legend.position = "none")
ARKO_plot_age
```

```{r}
ARKO_methylation <- cowplot::plot_grid(ARKO_plot_full, ARKO_plot_age, nrow=2, labels=c("A", "B"))
ARKO_methylation

ggsave("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Revision/FinalFigures/FigureSX_ARKOMethylation_2.png", ARKO_methylation, width=8,height=10, dpi=1000)
```


# t tests for all comparisons
```{r}
ARKO_MaleKO_MaleKOT <- ARKOdata %>% filter(FullDetailsBrief %in% c("MaleKO", "MaleKO+T"))
ARKO_MaleKO_FemaleWT <- ARKOdata %>% filter(FullDetailsBrief %in% c("MaleKO", "FemaleWT"))
ARKO_MaleKO_MaleWT <- ARKOdata %>% filter(FullDetailsBrief %in% c("MaleKO", "MaleWT"))
ARKO_MaleKOT_FemaleWT <- ARKOdata %>% filter(FullDetailsBrief %in% c("MaleKO+T", "FemaleWT"))
ARKO_MaleKOT_MaleWT <- ARKOdata %>% filter(FullDetailsBrief %in% c("MaleKO+T", "MaleWT"))
ARKO_FemaleWT_MaleWT <- ARKOdata %>% filter(FullDetailsBrief %in% c("FemaleWT", "MaleWT"))
```

# Male KO vs. Male KO+T
```{r}
t.test(ARKO_MaleKO_MaleKOT$CG.methylation~ARKO_MaleKO_MaleKOT$FullDetailsBrief)
```

# Male KO vs. Female WT
```{r}
t.test(ARKO_MaleKO_FemaleWT$CG.methylation~ARKO_MaleKO_FemaleWT$FullDetailsBrief)
```

# Male KO vs. Male WT
```{r}
t.test(ARKO_MaleKO_MaleWT$CG.methylation~ARKO_MaleKO_MaleWT$FullDetailsBrief)
```

# Male KO+T vs. Female WT
```{r}
t.test(ARKO_MaleKOT_FemaleWT$CG.methylation~ARKO_MaleKOT_FemaleWT$FullDetailsBrief)
```

# Male KO+T vs. Male WT
```{r}
t.test(ARKO_MaleKOT_MaleWT$CG.methylation~ARKO_MaleKOT_MaleWT$FullDetailsBrief)
```

# Female WT vs. Male WT
```{r}
t.test(ARKO_FemaleWT_MaleWT$CG.methylation~ARKO_FemaleWT_MaleWT$FullDetailsBrief)
```


## now, apply the data to the androgen clock
# I need to add to my new data: 
- a column with the beta value in it headed "cg21524116" in my new data
- AgeYears (base this off the existing AgeDays column)
- YPMP values

```{r}
ARKOdata$cg21524116 <- ARKOdata$Beta_corrected
ARKOdata$AgeYears <- ARKOdata$AgeDays/365
ARKOdata$YPMP <- ARKOdata$AgeYears-0.115068
ARKOdata$YPMP[ARKOdata$Sex=="Female"] <- 0
ARKOdata$YPMP[ARKOdata$YPMP < 0] <- 0
```

##### Muscle #####
# Now, to apply it to the new data.

```{r}
ARKOdata$Model_AndroAge_Muscleclock <- predict(muscle_model, newdata = ARKOdata)

Metrics::mdae(actual=ARKOdata$YPMP, predicted=ARKOdata$Model_AndroAge_Muscleclock)
# 0.634
```

```{r}
ARKOdata$Residual <- ARKOdata$Model_AndroAge_Muscleclock-ARKOdata$YPMP
```


```{r}
ARKO_MaleKO <- ARKOdata %>% filter(FullDetailsBrief=="MaleKO")
ARKO_MaleKOT <- ARKOdata %>% filter(FullDetailsBrief=="MaleKO+T")
ARKO_MaleWT <- ARKOdata %>% filter(FullDetailsBrief=="MaleWT")
ARKO_FemaleWT <- ARKOdata %>% filter(FullDetailsBrief=="FemaleWT")
```


```{r}
Metrics::mdae(actual=ARKOdata$YPMP, predicted=ARKOdata$Model_AndroAge_Muscleclock)
# 0.634
```

```{r}
Metrics::mdae(actual=ARKO_MaleKO$YPMP, predicted=ARKO_MaleKO$Model_AndroAge_Muscleclock)
# 0.909
```

```{r}
Metrics::mdae(actual=ARKO_MaleKOT$YPMP, predicted=ARKO_MaleKOT$Model_AndroAge_Muscleclock)
# 0.565
```

```{r}
Metrics::mdae(actual=ARKO_MaleWT$YPMP, predicted=ARKO_MaleWT$Model_AndroAge_Muscleclock)
# 1.10
```

```{r}
Metrics::mdae(actual=ARKO_FemaleWT$YPMP, predicted=ARKO_FemaleWT$Model_AndroAge_Muscleclock)
# 0.27
```

```{r}
ARKO_KWtest <- kruskal.test(Model_AndroAge_Muscleclock ~ FullDetailsBrief, data = ARKOdata)

if (ARKO_KWtest$p.value < 0.05) {
    ARKO_Dunn <- dunnTest(Model_AndroAge_Muscleclock ~ FullDetailsBrief, data = ARKOdata, method = "bh")
    print(ARKO_Dunn)
}
```



#### plotting
```{r}
d2_ARKO_aa <- ARKOdata %>%
  dplyr::group_by(FullDetailsBrief) %>%
  dplyr::summarise(
    mean = mean(Model_AndroAge_Muscleclock, na.rm = TRUE),
    sd = sd(Model_AndroAge_Muscleclock, na.rm = TRUE)
  )
```

```{r}
ARKOAgeAccel_anno2 <- ggplot(d2_ARKO_aa, aes(x=FullDetailsBrief, y=mean)) + 
  geom_bar(stat = "identity", fill="lightgrey", colour="black") + 
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.5) + 
  geom_point(ARKOdata, mapping = aes(x=FullDetailsBrief, y=Model_AndroAge_Muscleclock, colour=FullDetailsBrief), size=4, alpha=1/2) +
  ylab("AndroAge (Years)") + theme_classic() + geom_hline(yintercept=0) + 
  theme_light() +
  ggtitle("ARKO Muscle") +
  xlab("Treatment group") +
  scale_colour_manual(values = get_pal("Takahe")[c(2,1,4,3)]) +
  theme(legend.position="none") +
  geom_signif(y_position = c(2.6), xmin = c(1), xmax = c(4), tip_length = c(0.02, 0.02), annotation = c(""), textsize = 5) +
  geom_signif(y_position = c(2.35), xmin = c(1), xmax = c(3), tip_length = c(0.02, 0.02), annotation = c(""), textsize = 5) +
  geom_signif(y_position = c(2.1), xmin = c(1), xmax = c(2), tip_length = c(0.02, 0.02), annotation = c(""), textsize = 5) +
  geom_text(aes(x = 1.5, y = 2.15, label = "**"), size = 6, colour="black") +
  geom_text(aes(x = 2, y = 2.4, label = "***"), size = 6, colour="black") +
  geom_text(aes(x = 2.5, y = 2.65, label = "**"), size = 6, colour="black") +
  theme(text = element_text(size = 10), plot.title = element_text(hjust = 0.5)) +
  ylim(-1,2.7)

ARKOAgeAccel_anno2

ggsave("/Users/victoriasugrue/Dropbox (Otago University)/ARKOAgeAccel_anno2_fullannotations.png", ARKOAgeAccel_anno2, width=6,height=6, dpi=1000)
```


# Prenatal exposure mice

```{r}
prenatal_plot <- ggplot(prenatalDHTdata, aes(x=Treatment, y=CG.methylation, colour=Treatment)) +
  geom_boxplot(varwidth=T) + 
  geom_point(size=5, alpha=4/6) +
  labs(title="PNA mice", 
       x="Treatment",
       y="Methylation (%)") +
  theme_light() +
  theme(legend.position = "none") + 
  #theme(text = element_text(size = 15)) +
  ylim(55,85) +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)])

prenatal_plot

ggsave("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Revision/FinalFigures/FigureSX_PNAMethylation.png", prenatal_plot, width=4,height=3, dpi=1000)
```

```{r}
t.test(prenatalDHTdata$CG.methylation~prenatalDHTdata$Treatment)
```


```{r}
prenatalDHTdata$cg21524116 <- prenatalDHTdata$Beta_corrected
prenatalDHTdata$AgeYears <- prenatalDHTdata$AgeDays/365
prenatalDHTdata$YPMP <- prenatalDHTdata$AgeYears-0.115068
prenatalDHTdata$Sex <- "Female"
prenatalDHTdata$YPMP[prenatalDHTdata$Sex=="Female"] <- 0
prenatalDHTdata$YPMP[prenatalDHTdata$YPMP < 0] <- 0
prenatalDHTdata <- prenatalDHTdata %>%
  mutate(Treatment = ifelse(Treatment == "DHT", "PNA", Treatment))


prenatalDHTdata$Model_AndroAge_Muscleclock <- predict(muscle_model, newdata = prenatalDHTdata)

Metrics::mdae(actual=prenatalDHTdata$YPMP, predicted=prenatalDHTdata$Model_AndroAge_Muscleclock)
# 0.24
```

```{r}
prenatalDHTdata$AndroAgeAccel <- prenatalDHTdata$Model_AndroAge_Muscleclock-prenatalDHTdata$YPMP
```


```{r}
prenatalDHT <- prenatalDHTdata %>% filter(Treatment=="PNA") 
prenatalControl <- prenatalDHTdata %>% filter(Treatment=="Control")
```


```{r}
Metrics::mdae(actual=prenatalDHT$YPMP, predicted=prenatalDHT$Model_AndroAge_Muscleclock)
# 0.241
```

```{r}
Metrics::mdae(actual=prenatalControl$YPMP, predicted=prenatalControl$Model_AndroAge_Muscleclock)
# 0.247
```

```{r}
prenatalDHTdata$Residual <- prenatalDHTdata$Model_AndroAge_Muscleclock-prenatalDHTdata$YPMP

wilcox.test(prenatalDHTdata$Model_AndroAge_Muscleclock~prenatalDHTdata$Treatment)

wilcox.test(prenatalDHTdata$Residual~prenatalDHTdata$Treatment)
```

```{r}
d2prenatal <- prenatalDHTdata %>%
  dplyr::group_by(Treatment) %>%
  dplyr::summarise(
    mean = mean(Model_AndroAge_Muscleclock, na.rm = TRUE),
    sd = sd(Model_AndroAge_Muscleclock, na.rm = TRUE)
  )
```

```{r}
AndroAgePrenatal <- ggplot(d2prenatal, aes(x=Treatment, y=mean)) + 
  geom_bar(stat = "identity", fill="lightgrey", colour="black") + 
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.5) + 
  geom_point(prenatalDHTdata, mapping = aes(x=Treatment, y=prenatalDHTdata$Model_AndroAge_Muscleclock, colour=Treatment), size=4, alpha=1/2) +
  ylab("AndroAge (Years)") + theme_classic() + geom_hline(yintercept=0) + 
  theme_light() +
  ggtitle("Prenatal DHT Muscle") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme(legend.position="none", plot.title = element_text(hjust = 0.5)) +
  theme(text = element_text(size = 10)) +
  ylim(-1,2.7)
AndroAgePrenatal
```


## plot all things together

```{r}
v1 <- cowplot::plot_grid(AndroAgeTail, AndroAgeMuscle, AndroAgePrenatal, nrow=1, labels=c("B", "C", "D"))
v1

v2 <- cowplot::plot_grid(DHT_combined, v1, nrow=2, labels=c("A"))
v2

```


# Plot together 
```{r}
#top <- cowplot::plot_grid(DHT_combined, AndroAgeTail, AndroAgeMuscle, nrow=1, rel_widths = c(1, 0.4, 0.4))
#top

#bottom <- cowplot::plot_grid(AndroAgePrenatal, ARKOAgeAccel_anno2, nrow=1, rel_widths = c(0.3, 1))
#bottom

#both <- cowplot::plot_grid(top, bottom, nrow=2, rel_heights = c(1,1))
#both

#ggsave("/Users/victoriasugrue/Dropbox (Otago University)/checkfig4layout_v4_corrected.png", both, width=12,height=8, dpi=1000)
#ggsave("/Users/victoriasugrue/Dropbox (Otago University)/checkfig4layout_v4.1_corrected.png", both, width=6,height=8, dpi=1000)

```

# all plot codes to edit 
```{r}
DHT_combined2 <- ggplot(capsuleDHTdata, aes(x=Tissue, y=CG.methylation, colour=Treatment)) +
  geom_boxplot(varwidth=T) + 
  geom_point(size=5, alpha=4/6, position=position_dodge(width=0.75)) +
  labs(title="  ", 
       x="Tissue",
       y="Methylation (%)") +
  theme_light() +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  ylim(30,100) +
  geom_signif(y_position = c(95), xmin = c(0.75), xmax = c(1.25), tip_length = c(0.02, 0.02), annotation = c(""), textsize = 5, colour="black") +
  geom_signif(y_position = c(95), xmin = c(1.75), xmax = c(2.25), tip_length = c(0.02, 0.02), annotation = c(""), textsize = 5, colour="black") +
  geom_signif(y_position = c(95), xmin = c(2.75), xmax = c(3.25), tip_length = c(0.02, 0.02), annotation = c(""), textsize = 5, colour="black") +
  geom_signif(y_position = c(95), xmin = c(3.75), xmax = c(4.25), tip_length = c(0.02, 0.02), annotation = c(""), textsize = 5, colour="black") +
  geom_text(aes(x = 1, y = 96.5, label = "***"), size = 6, colour="black") +
  geom_text(aes(x = 2, y = 96.5, label = "***"), size = 6, colour="black") +
  geom_text(aes(x = 3, y = 96.5, label = "**"), size = 6, colour="black") +
  geom_text(aes(x = 4, y = 96.5, label = "***"), size = 6, colour="black") 

DHT_combined2
```

```{r}
AndroAgeTail <- ggplot(d2tail, aes(x=Treatment, y=mean)) + 
  geom_bar(stat = "identity", fill="lightgrey", colour="black") + 
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.5) + 
  geom_point(capsule_tail, mapping = aes(x=Treatment, y=Model_AndroAge_Tailclock, colour=Treatment), size=4, alpha=1/2) +
  ylab("AndroAge (Years)") + theme_classic() + geom_hline(yintercept=0) + 
  theme_light() +
  ggtitle("C-DHT Tail") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme(legend.position="none", plot.title = element_text(hjust = 0.5))+
  theme(text = element_text(size = 10)) +
  geom_signif(y_position = c(2.6), xmin = c(1), xmax = c(2), tip_length = c(0.02, 0.02), annotation = c(""), textsize = 5) +
  geom_text(aes(x = 1.5, y = 2.65, label = "***"), size = 6, colour="black") +
  ylim(-1, 2.7)
AndroAgeTail
```

```{r}
AndroAgeMuscle <- ggplot(d2muscle, aes(x=Treatment, y=mean)) + 
  geom_bar(stat = "identity", fill="lightgrey", colour="black") + 
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.5) + 
  geom_point(capsule_muscle, mapping = aes(x=Treatment, y=Model_AndroAge_Muscleclock, colour=Treatment), size=4, alpha=1/2) +
  ylab("AndroAge (Years)") + theme_classic() + geom_hline(yintercept=0) + 
  theme_light() +
  ggtitle("C-DHT Muscle") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme(legend.position="none", plot.title = element_text(hjust = 0.5)) +
  theme(text = element_text(size = 10)) +
  geom_signif(y_position = c(2.6), xmin = c(1), xmax = c(2), tip_length = c(0.02, 0.02), annotation = c(""), textsize = 5) +
  geom_text(aes(x = 1.5, y = 2.65, label = "***"), size = 6, colour="black") +
  ylim(-1, 2.7)
AndroAgeMuscle
```

```{r}
AndroAgePrenatal <- ggplot(d2prenatal, aes(x=Treatment, y=mean)) + 
  geom_bar(stat = "identity", fill="lightgrey", colour="black") + 
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.5) + 
  geom_point(prenatalDHTdata, mapping = aes(x=Treatment, y=prenatalDHTdata$Model_AndroAge_Muscleclock, colour=Treatment), size=4, alpha=1/2) +
  ylab("AndroAge (Years)") + theme_classic() + geom_hline(yintercept=0) + 
  theme_light() +
  ggtitle("PNA Muscle") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme(legend.position="none", plot.title = element_text(hjust = 0.5)) +
  theme(text = element_text(size = 10)) +
  ylim(-1,2.7)
AndroAgePrenatal
```

```{r}
ARKOAgeAccel_anno2 <- ggplot(d2_ARKO_aa, aes(x=FullDetailsBrief, y=mean)) + 
  geom_bar(stat = "identity", fill="lightgrey", colour="black") + 
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.5) + 
  geom_point(ARKOdata, mapping = aes(x=FullDetailsBrief, y=Model_AndroAge_Muscleclock, colour=FullDetailsBrief), size=4, alpha=1/2) +
  ylab("AndroAge (Years)") + theme_classic() + geom_hline(yintercept=0) + 
  theme_light() +
  ggtitle("ARKO Muscle") +
  xlab("Treatment group") +
  scale_colour_manual(values = get_pal("Takahe")[c(2,1,4,3)]) +
  theme(legend.position="none") +
  geom_signif(y_position = c(2.6), xmin = c(1), xmax = c(4), tip_length = c(0.02, 0.02), annotation = c(""), textsize = 5) +
  geom_signif(y_position = c(2.35), xmin = c(1), xmax = c(3), tip_length = c(0.02, 0.02), annotation = c(""), textsize = 5) +
  geom_signif(y_position = c(2.1), xmin = c(1), xmax = c(2), tip_length = c(0.02, 0.02), annotation = c(""), textsize = 5) +
  geom_text(aes(x = 1.5, y = 2.15, label = "**"), size = 6, colour="black") +
  geom_text(aes(x = 2, y = 2.4, label = "***"), size = 6, colour="black") +
  geom_text(aes(x = 2.5, y = 2.65, label = "**"), size = 6, colour="black") +
  theme(text = element_text(size = 10), plot.title = element_text(hjust = 0.5)) +
  ylim(-1,2.7)

ARKOAgeAccel_anno2
```




```{r}
bottom2 <- cowplot::plot_grid(AndroAgeTail, AndroAgeMuscle, AndroAgePrenatal, ARKOAgeAccel_anno2, nrow=1, rel_widths=c(0.5, 0.5, 0.5, 0.8), labels=c("B", "C", "D", "E"))
bottom2

figure4 <- cowplot::plot_grid(DHT_combined2, bottom2, nrow=2, rel_heights = c(0.8, 1), labels=c("A"))
figure4

ggsave("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Revision/FinalFigures/Figure4.png", figure4, width=10,height=8, dpi=1000)
```
