---
title: "Mouse Androgen Clock Construction"
author: "Victoria Sugrue"
date: 'Last updated: 28 July 2023'
output:
  word_document:
    toc: no
    toc_depth: '6'
  html_document:
    toc: no
    toc_float: no
    toc_depth: 6
  pdf_document:
    toc: no
    toc_depth: 5
    df_print: kable
    latex_engine: xelatex
subtitle: ' '
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load required packages
```{r}
library(ggplot2)
library(caret)
library(dplyr)
library(Manu)
library(knitr)
library(lmtest)
library(ggfortify)
```

# Data preparation
```{r}
mouse_data <- read.csv("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Data/Mouse_ArrayData.csv")
mouse_data <- mouse_data %>% filter(Tissue=="Muscle" | Tissue=="Tail")

# Setting AndroAge with predicted time of puberty onset = age minus 6 weeks, adjust to 0 where negative
mouse_data$YPMP <- mouse_data$Age-0.115068
mouse_data$YPMP[mouse_data$Sex=="F"] <- 0
mouse_data$YPMP[mouse_data$YPMP < 0] <- 0

# Change sex labels
mouse_data$Sex[mouse_data$Sex=="F"] <- "Female"
mouse_data$Sex[mouse_data$Sex=="M"] <- "Male"

# Subset by tissue
mouse_muscle <- mouse_data %>% filter(Tissue=="Muscle")
mouse_tail <- mouse_data %>% filter(Tissue=="Tail")
```


# A birds-eye view of the data 

```{r}
# Age distribution
age_dist <- ggplot(mouse_data, aes(x = Age)) +
  geom_histogram(aes(fill = Tissue), position = "dodge", alpha = 1, bins = 15) +
  scale_fill_manual(values = get_pal("Takahe")) +
  theme_light() +
  facet_grid(~Sex) +
  labs(title = "Distribution of age by sex (Mouse, array)", fill = "Sex", x="Age (Years)") +
  ylab("Frequency")
age_dist
```

```{r}
#or, by tissue
age_dist_tis <- ggplot(mouse_data, aes(x = Age)) +
  geom_histogram(aes(fill = Sex), position = "dodge", alpha = 1, bins = 15) +
  scale_fill_manual(values = get_pal("Takahe")[c(1,2,4)]) +
  theme_light() +
  facet_wrap(~Tissue, ncol=3, scales="free") +
  labs(title = "Distribution of age by sex (Mouse, array)", fill = "Sex", x="Age (Years)") +
  ylab("Frequency")
age_dist_tis
```

# Plotting methylation data

```{r}
MKLN1_mice <- ggplot(mouse_data, aes(x=Age, y=cg21524116*100, colour=Sex)) +
  geom_jitter(width=0.05, size=2.5, alpha=2/3) +
  geom_smooth(method="lm") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2,4)]) +
  theme_light() +
  facet_wrap(~Tissue, ncol=3, scales="free") +
  labs(colour="Sex", y="Methylation (%)", x="Age (Years)")
MKLN1_mice
```

Testing both tissues

##### Tail #####

# Fit androgen clock to mouse tail
```{r}
male_tail <- mouse_tail %>% filter(Sex=="Male")

tail_model <- lm(YPMP ~ cg21524116, data=male_tail) 
summary(tail_model)
```

# Diagnostic plots

```{r}
autoplot(tail_model, which=c(1,2,3,5), size=3, alpha=4/5, colour="#313657", smooth.colour="#DD3C51", smooth.linetype="dashed", label.size = 3) + theme_light() + theme(plot.title = element_text(color = "white"))
```

```{r}
fitted_tail <- as.data.frame(tail_model$fitted.values)
fitted_tail["YPMP"] <- male_tail$YPMP
fitted_tail$YPMP <- as.numeric(fitted_tail$YPMP)
fitted_tail$Fit_YPMP <- as.numeric(fitted_tail$`tail_model$fitted.values`)
```

# Calculating error
## All

```{r}
MAE_tail_model <- as.data.frame(Metrics::mdae(actual=fitted_tail$YPMP, predicted=fitted_tail$`tail_model$fitted.values`))
MAE_tail_model # 0.208
```

```{r}
cor(fitted_tail$YPMP, fitted_tail$Fit_YPMP, method="pearson")

tail_modelfit <- ggplot(fitted_tail, aes(x=YPMP, y=Fit_YPMP)) +
  geom_point(size=4, alpha=3/4, colour="#313657") + 
  geom_smooth(method="lm", se=F, colour="#DD3C51") +
  geom_abline(linetype="dashed") +
  theme_light() +
  ylab("Model AndroAge (Years)") +
  xlab("Years Post Male Puberty (YPMP)") +
    labs(title = "Tail",
              subtitle = "MAE=0.208, cor=0.811") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5))
tail_modelfit
```

# Breusch-Pagan test of variance (homo/heteroscedasticity)

```{r}
plot(tail_model$residuals)
bptest(tail_model)
```

# p value >0.05 suggests that we have homoscedasticity. good! 

# Shapiro-Wilk test of normality

```{r}
shapiro.test(tail_model$residuals)
```

# p value <0.05 suggests data is normally distributed.  good! (just)

# Leave-one-out cross validation

```{r}
male_tail$cg21524116 <- as.numeric(male_tail$cg21524116)
male_tail$YPMP <- as.numeric(male_tail$YPMP)
ctrl <- trainControl(method="LOOCV")
tail_modelLOO <- train(YPMP ~ cg21524116, data=male_tail, method="lm", trControl=ctrl)
summary(tail_modelLOO)
```

```{r}
print(tail_modelLOO)
```

```{r}
fittedLOO <- as.data.frame(tail_modelLOO$pred)
fittedLOO["YPMP"] <- male_tail$YPMP

tail_modelLOO_MAE <- as.data.frame(Metrics::mdae(actual=fittedLOO$YPMP, predicted=fittedLOO$pred))
tail_modelLOO_MAE # 0.237
```


# Now, to include females.
# First, apply the formula of the male androgen clock to this group. What happens? 

```{r}
female_tail <- mouse_tail %>% filter(Sex=="Female")
female_tail$Model_AndroAge <- predict(tail_model, newdata = female_tail)

female_MAE_tail_model <- as.data.frame(Metrics::mdae(actual=female_tail$YPMP, predicted=female_tail$Model_AndroAge))
female_MAE_tail_model
# Female MAE = 0.103
```

# LOOCV across both sexes


```{r}
mouse_tail$cg21524116 <- as.numeric(mouse_tail$cg21524116)
mouse_tail$YPMP <- as.numeric(mouse_tail$YPMP)

predicted_YPMP_mouse_tail_6 <- predict(tail_modelLOO, newdata = mouse_tail)
median(abs(predicted_YPMP_mouse_tail_6 - mouse_tail$YPMP))
# 0.120
```






# Fit androgen clock to mouse muscle
```{r}
male_muscle <- mouse_muscle %>% filter(Sex=="Male")

muscle_model <- lm(YPMP ~ cg21524116, data=male_muscle) 
summary(muscle_model)
```

# Diagnostic plots

```{r}
autoplot(muscle_model, which=c(1,2,3,5), size=3, alpha=4/5, colour="#313657", smooth.colour="#DD3C51", smooth.linetype="dashed", label.size = 3) + theme_light() + theme(plot.title = element_text(color = "white"))
```

```{r}
fitted_muscle <- as.data.frame(muscle_model$fitted.values)
fitted_muscle["YPMP"] <- male_muscle$YPMP
fitted_muscle$YPMP <- as.numeric(fitted_muscle$YPMP)
fitted_muscle$Fit_YPMP <- as.numeric(fitted_muscle$`muscle_model$fitted.values`)
```

# Calculating error
## All

```{r}
MAE_muscle_model <- as.data.frame(Metrics::mdae(actual=fitted_muscle$YPMP, predicted=fitted_muscle$`muscle_model$fitted.values`))
MAE_muscle_model # 0.457
```

```{r}
cor(fitted_muscle$YPMP, fitted_muscle$Fit_YPMP, method="pearson") #0.491


muscle_modelfit <- ggplot(fitted_muscle, aes(x=YPMP, y=Fit_YPMP)) +
  geom_point(size=4, alpha=3/4, colour="#313657") + 
  geom_smooth(method="lm", se=F, colour="#DD3C51") +
  geom_abline(linetype="dashed") +
  theme_light() +
  ylab("Model AndroAge (Years)") +
  xlab("Years Post Male Puberty (YPMP)") +
    labs(title = "Muscle",
              subtitle = "MAE=0.457, cor=0.491") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5))
muscle_modelfit

ggsave("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Figures/muscle_modelfit.png",muscle_modelfit, width= 5, height = 5, dpi=1000)

```


# Breusch-Pagan test of variance (homo/heteroscedasticity)

```{r}
plot(muscle_model$residuals)
bptest(muscle_model)
```

# p value >0.05 suggests that we have homoscedasticity. good! 

# Shapiro-Wilk test of normality

```{r}
shapiro.test(muscle_model$residuals)
```

# p value <0.05 suggests data isn't normally distributed. not good!

# Leave-one-out cross validation

```{r}
male_muscle$cg21524116 <- as.numeric(male_muscle$cg21524116)
male_muscle$YPMP <- as.numeric(male_muscle$YPMP)
ctrl <- trainControl(method="LOOCV")
muscle_modelLOO <- train(YPMP ~ cg21524116, data=male_muscle, method="lm", trControl=ctrl)
summary(muscle_modelLOO)
```

```{r}
print(muscle_modelLOO)
```

```{r}
fittedLOO <- as.data.frame(muscle_modelLOO$pred)
fittedLOO["YPMP"] <- male_muscle$YPMP

muscle_modelLOO_MAE <- as.data.frame(Metrics::mdae(actual=fittedLOO$YPMP, predicted=fittedLOO$pred))
muscle_modelLOO_MAE # 0.530
```


# Now, to include females.
# First, apply the formula of the male androgen clock to this group. What happens? 

```{r}
female_muscle <- mouse_muscle %>% filter(Sex=="Female")
female_muscle$Model_AndroAge <- predict(muscle_model, newdata = female_muscle)

female_MAE_muscle_model <- as.data.frame(Metrics::mdae(actual=female_muscle$YPMP, predicted=female_muscle$Model_AndroAge))
female_MAE_muscle_model
# Female MAE = 0.256
```

# LOOCV across both sexes

```{r}
mouse_muscle$cg21524116 <- as.numeric(mouse_muscle$cg21524116)
mouse_muscle$YPMP <- as.numeric(mouse_muscle$YPMP)

predicted_YPMP_mouse_muscle_8 <- predict(muscle_modelLOO, newdata = mouse_muscle)
median(abs(predicted_YPMP_mouse_muscle_8 - mouse_muscle$YPMP))
#0.343
```








# Selected clock: Tail

```{r}
mouse_tail$Model_AndroAge <- predict(tail_model, newdata = mouse_tail)

MAE_tail_model_all <- as.data.frame(Metrics::mdae(actual=mouse_tail$YPMP, predicted=mouse_tail$Model_AndroAge))
MAE_tail_model_all
#0.120
```


```{r}
mouse_tail$LOO_AndroAge <- predicted_YPMP_mouse_tail_6
```


```{r}
cor(mouse_tail$YPMP, mouse_tail$LOO_AndroAge, method="pearson") #0.861

obs_exp_mousetail_model <- ggplot(mouse_tail, aes(x=YPMP, y=LOO_AndroAge, colour=Sex)) +
 geom_point(size=4, alpha=3/4) + 
  geom_smooth(method="lm", se=F, colour="#1F6683") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme_light() +
  geom_abline(linetype="dashed") +
  ylab("Model AndroAge (Years)") +
  xlab("Years Post Male Puberty (YPMP)") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5)) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(title.position = "top", title.hjust=0.5))

obs_exp_mousetail_model

ggsave("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Figures/Obs_Exp_AndroAge_mousetail_model2_6weeks.png",obs_exp_mousetail_model, width= 3.7, height = 4.2, dpi=1000)
```


```{r}
write.csv(mouse_tail, "/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Data/mouse_data_modelled_tail_model_6weeks.csv")
```



# Applying the model to the DHT mouse data
```{r}
DHTdata <- read.csv("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Data/DHTmice.csv")
DHTdata$cg21524116 <- DHTdata$Methylation/100

DHTtail <- DHTdata %>% filter(Tissue=="Tail")

DHTtail$Model_AndroAge <- predict(tail_model, newdata = DHTtail)

MAE_DHTtail_all <- as.data.frame(Metrics::mdae(actual=0, predicted=DHTtail$Model_AndroAge))
MAE_DHTtail_all
# 0.755
```
## controls only
```{r}
DHTtail_control <- DHTtail %>% filter(Treatment=="Blank")

DHTtail_control$Model_AndroAge <- predict(tail_model, newdata = DHTtail_control)

MAE_DHTtail_control <- as.data.frame(Metrics::mdae(actual=0, predicted=DHTtail_control$Model_AndroAge))
MAE_DHTtail_control
# 0.576
```

## DHT only
```{r}
DHTtail_DHT <- DHTtail %>% filter(Treatment=="DHT")

DHTtail_DHT$Model_AndroAge <- predict(tail_model, newdata = DHTtail_DHT)

MAE_DHTtail_DHT <- as.data.frame(Metrics::mdae(actual=0, predicted=DHTtail_DHT$Model_AndroAge))
MAE_DHTtail_DHT
# 2.259
```







