---
title: "Mouse Androgen Clock Construction"
author: "Victoria Sugrue"
date: 'Last updated: 26 July 2024'
output:
  word_document:
    toc: no
    toc_depth: '6'
  html_document:
    toc: no
    toc_float: no
    toc_depth: 6
  pdf_document:
    toc: no
    toc_depth: 5
    df_print: kable
    latex_engine: xelatex
subtitle: ' '
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load required packages
```{r}
library(ggplot2)
library(caret)
library(dplyr)
library(Manu)
library(knitr)
library(lmtest)
library(ggfortify)
library(cowplot)
```

# Data preparation
```{r}
mouse_data <- read.csv("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Data/Mouse_ArrayData_Full_TailMuscle.csv")
mouse_data <- mouse_data %>% filter(Tissue=="Muscle" | Tissue=="Tail")

# Setting AndroAge with predicted time of puberty onset = age minus 6 weeks, adjust to 0 where negative
mouse_data$YPMP <- mouse_data$Age-0.115068
mouse_data$YPMP[mouse_data$Sex=="F"] <- 0
mouse_data$YPMP[mouse_data$YPMP < 0] <- 0

# Change sex labels
mouse_data$Sex[mouse_data$Sex=="F"] <- "Female"
mouse_data$Sex[mouse_data$Sex=="M"] <- "Male"

# Subset by tissue
mouse_muscle <- mouse_data %>% filter(Tissue=="Muscle")
mouse_tail <- mouse_data %>% filter(Tissue=="Tail")

# Subset by sex
male_muscle <- mouse_muscle %>% filter(Sex=="Male")
female_muscle <- mouse_muscle %>% filter(Sex=="Female")
male_tail <- mouse_tail %>% filter(Sex=="Male")
female_tail <- mouse_tail %>% filter(Sex=="Female")
```


# A birds-eye view of the data 

```{r}
# Age distribution
age_dist <- ggplot(mouse_data, aes(x = Age)) +
  geom_histogram(aes(fill = Tissue), position = "dodge", alpha = 1, bins = 15) +
  scale_fill_manual(values = get_pal("Takahe")) +
  theme_light() +
  facet_grid(~Sex) +
  labs(title = "Distribution of age by sex (Mouse, array)", fill = "Sex", x="Age (Years)") +
  ylab("Frequency")
age_dist
```

```{r}
#or, by tissue
age_dist_tis <- ggplot(mouse_data, aes(x = Age)) +
  geom_histogram(aes(fill = Sex), position = "dodge", alpha = 1, bins = 15) +
  scale_fill_manual(values = get_pal("Takahe")[c(1,2,4)]) +
  theme_light() +
  facet_wrap(~Tissue, ncol=3, scales="free") +
  labs(title = "Distribution of age by sex (Mouse, array)", fill = "Sex", x="Age (Years)") +
  ylab("Frequency")
age_dist_tis
```

# Plotting methylation data

```{r}
MKLN1_mice <- ggplot(mouse_data, aes(x=Age, y=cg21524116*100, colour=Sex)) +
  geom_jitter(width=0.05, size=2.5, alpha=2/3) +
  geom_smooth(method="lm") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2,4)]) +
  theme_light() +
  facet_wrap(~Tissue, ncol=3, scales="free") +
  labs(colour="Sex", y="Methylation (%)", x="Age (Years)")
MKLN1_mice
```

Testing both tissues

##### Tail #####

# Fit androgen clock to mouse tail
```{r}
tail_model <- lm(YPMP ~ cg21524116, data=male_tail) 
summary(tail_model)
```

# Diagnostic plots

```{r}
autoplot(tail_model, which=c(1,2,3,5), size=3, alpha=4/5, colour="#313657", smooth.colour="#DD3C51", smooth.linetype="dashed", label.size = 3) + theme_light() + theme(plot.title = element_text(color = "white"))
```

```{r}
fitted_tail <- as.data.frame(tail_model$fitted.values)
colnames(fitted_tail)[1] <- "Predicted_AndroAge"
fitted_tail["YPMP"] <- male_tail$YPMP
fitted_tail$YPMP <- as.numeric(fitted_tail$YPMP)
```

# Calculating error
## All

```{r}
Metrics::mdae(actual=fitted_tail$YPMP, predicted=fitted_tail$Predicted_AndroAge)
# 0.117

cor.test(fitted_tail$YPMP, fitted_tail$Predicted_AndroAge, method="pearson")
# 0.802
```

```{r}
tail_modelfit <- ggplot(fitted_tail, aes(x=YPMP, y=Predicted_AndroAge)) +
  geom_point(size=4, alpha=3/4, colour="#313657") + 
  geom_smooth(method="lm", se=F, colour="#DD3C51") +
  geom_abline(linetype="dashed") +
  theme_light() +
  ylab("Model AndroAge (Years)") +
  xlab("Years Post Male Puberty (YPMP)") +
    labs(title = "Tail",
              subtitle = "MAE=0.117, cor=0.802") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5))
tail_modelfit
```

# Leave-one-out cross validation

```{r}
male_tail$cg21524116 <- as.numeric(male_tail$cg21524116)
male_tail$YPMP <- as.numeric(male_tail$YPMP)
ctrl <- trainControl(method="LOOCV", savePredictions = TRUE)
tail_modelLOO <- train(YPMP ~ cg21524116, data=male_tail, method="lm", trControl=ctrl)
summary(tail_modelLOO)
print(tail_modelLOO)
```

```{r}
fitted_LOO_tail <- as.data.frame(tail_modelLOO$pred)
colnames(fitted_LOO_tail)[1] <- "Predicted_AndroAge_LOO"
colnames(fitted_LOO_tail)[2] <- "YPMP"

male_tail$Predicted_AndroAge_LOO <- fitted_LOO_tail$Predicted_AndroAge_LOO

Metrics::mdae(actual=male_tail$Predicted_AndroAge_LOO, predicted=male_tail$YPMP)
# 0.123

cor.test(male_tail$YPMP, male_tail$Predicted_AndroAge_LOO, method="pearson")
# 0.782
```

# Apply to whole dataset
```{r}
mouse_tail$Predicted_AndroAge <- predict(tail_model, newdata = mouse_tail)

Metrics::mdae(actual=mouse_tail$Predicted_AndroAge, predicted=mouse_tail$YPMP) # 0.114
cor.test(mouse_tail$YPMP, mouse_tail$Predicted_AndroAge, method="pearson") # 0.814
```


# For each sex independently
```{r}
male_tail <- mouse_tail %>% filter(Sex=="Male") 
Metrics::mdae(actual=male_tail$Predicted_AndroAge, predicted=male_tail$YPMP) # 0.117
cor.test(male_tail$YPMP, male_tail$Predicted_AndroAge, method="pearson") # 0.802

female_tail <- mouse_tail %>% filter(Sex=="Female")
Metrics::mdae(actual=female_tail$Predicted_AndroAge, predicted=female_tail$YPMP) # 0.113
```

```{r}
obs_exp_mousetail_model <- ggplot(mouse_tail, aes(x=YPMP, y=Predicted_AndroAge, colour=Sex)) +
 geom_point(size=4, alpha=3/4) + 
  geom_smooth(method="lm", se=F, colour="#1F6683") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme_light() +
  geom_abline(linetype="dashed") +
  ylab("Model AndroAge (Years)") +
  xlab("Years Post Male Puberty (YPMP)") +
  labs(title="Tail", subtitle="MAE=0.114, cor=0.814") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5)) +
  theme(legend.position = "none") 
obs_exp_mousetail_model
```






##### Muscle #####

# Fit androgen clock to mouse muscle
```{r}
muscle_model <- lm(YPMP ~ cg21524116, data=male_muscle) 
summary(muscle_model)
```

# Diagnostic plots

```{r}
autoplot(muscle_model, which=c(1,2,3,5), size=3, alpha=4/5, colour="#313657", smooth.colour="#DD3C51", smooth.linetype="dashed", label.size = 3) + theme_light() + theme(plot.title = element_text(color = "white"))
```

```{r}
fitted_muscle <- as.data.frame(muscle_model$fitted.values)
colnames(fitted_muscle)[1] <- "Predicted_AndroAge"
fitted_muscle["YPMP"] <- male_muscle$YPMP
fitted_muscle$YPMP <- as.numeric(fitted_muscle$YPMP)
```

# Calculating error
## All

```{r}
Metrics::mdae(actual=fitted_muscle$YPMP, predicted=fitted_muscle$Predicted_AndroAge)
# 0.488

cor.test(fitted_muscle$YPMP, fitted_muscle$Predicted_AndroAge, method="pearson")
# 0.634
```

```{r}
muscle_modelfit <- ggplot(fitted_muscle, aes(x=YPMP, y=Predicted_AndroAge)) +
  geom_point(size=4, alpha=3/4, colour="#313657") + 
  geom_smooth(method="lm", se=F, colour="#DD3C51") +
  geom_abline(linetype="dashed") +
  theme_light() +
  ylab("Model AndroAge (Years)") +
  xlab("Years Post Male Puberty (YPMP)") +
    labs(title = "Muscle",
              subtitle = "MAE=0.488, cor=0.634") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5))
muscle_modelfit
```

# Leave-one-out cross validation

```{r}
male_muscle$cg21524116 <- as.numeric(male_muscle$cg21524116)
male_muscle$YPMP <- as.numeric(male_muscle$YPMP)
ctrl <- trainControl(method="LOOCV", savePredictions = TRUE)
muscle_modelLOO <- train(YPMP ~ cg21524116, data=male_muscle, method="lm", trControl=ctrl)
summary(muscle_modelLOO)
print(muscle_modelLOO)
```

```{r}
fitted_LOO_muscle <- as.data.frame(muscle_modelLOO$pred)
colnames(fitted_LOO_muscle)[1] <- "Predicted_AndroAge_LOO"
colnames(fitted_LOO_muscle)[2] <- "YPMP"

male_muscle$Predicted_AndroAge_LOO <- fitted_LOO_muscle$Predicted_AndroAge_LOO

Metrics::mdae(actual=male_muscle$Predicted_AndroAge_LOO, predicted=male_muscle$YPMP)
# 0.500

cor.test(male_muscle$YPMP, male_muscle$Predicted_AndroAge_LOO, method="pearson")
# 0.626
```

# Apply to whole dataset
```{r}
mouse_muscle$Predicted_AndroAge <- predict(muscle_model, newdata = mouse_muscle)

Metrics::mdae(actual=mouse_muscle$Predicted_AndroAge, predicted=mouse_muscle$YPMP) # 0.419
cor.test(mouse_muscle$YPMP, mouse_muscle$Predicted_AndroAge, method="pearson") # 0.754
```


# For each sex independently
```{r}
male_muscle <- mouse_muscle %>% filter(Sex=="Male") 
Metrics::mdae(actual=male_muscle$Predicted_AndroAge, predicted=male_muscle$YPMP) # 0.488
cor.test(male_muscle$YPMP, male_muscle$Predicted_AndroAge, method="pearson") # 0.634

female_muscle <- mouse_muscle %>% filter(Sex=="Female")
Metrics::mdae(actual=female_muscle$Predicted_AndroAge, predicted=female_muscle$YPMP) # 0.289
```



```{r}
obs_exp_mousemuscle_model <- ggplot(mouse_muscle, aes(x=YPMP, y=Predicted_AndroAge, colour=Sex)) +
 geom_point(size=4, alpha=3/4) + 
  geom_smooth(method="lm", se=F, colour="#1F6683") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2)]) +
  theme_light() +
  geom_abline(linetype="dashed") +
  ylab("Model AndroAge (Years)") +
  xlab("Years Post Male Puberty (YPMP)") +
  labs(title="Muscle", subtitle="MAE=0.419, cor=0.754") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5)) +
  theme(legend.position = "none") 
obs_exp_mousemuscle_model
```


# Plot together
```{r}
legend_b <- get_legend(
  obs_exp_mousetail_model + guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom"))

mousemodels1 <- cowplot::plot_grid(obs_exp_mousetail_model, obs_exp_mousemuscle_model, nrow=1, labels=c("B", "C"))

mousemodels <- cowplot::plot_grid(mousemodels1, legend_b, nrow=2, rel_heights = c(1,0.1))
mousemodels

mousemodels_cohortspace <- cowplot::plot_grid(NULL, mousemodels, ncol=1, rel_heights = c(0.85,1), labels=c("A"))

ggsave("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Revision/FinalFigures/Figure3_noCohort.png",mousemodels_cohortspace, width=8, height = 8, dpi=1000)
```


Plotting female model_AndroAge
```{r}
mouse_tail_female <- mouse_tail %>% filter(Sex=="Female") 

plot1 <- ggplot(mouse_tail_female, aes(x=Age, y=Predicted_AndroAge)) +
  geom_point(size=5, alpha=2/3) +
  ggtitle("Chronological Age vs modelled AndroAge, Female mice only") +
  theme_light() +
  ylab("AndroAge (Years)") +
  xlab("Chronological Age (Years)") +
  geom_hline(yintercept=0, linetype="dashed") +
  ylim(-1, 1)

plot1
```



