---
title: "Sheep Androgen Clock Construction - BBA-seq approach, 5% error"
author: "Victoria Sugrue"
date: 'Last updated: 26 July 2024'
output:
  word_document:
    toc: yes
    toc_depth: '6'
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 6
  pdf_document:
    toc: yes
    toc_depth: 5
    df_print: kable
    latex_engine: xelatex
subtitle: ' '
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load required packages
```{r}
library(ggplot2)
library(caret)
library(dplyr)
library(Manu)
library(knitr)
library(lmtest)
library(ggfortify)
```


### Data preparation
```{r}
sheep_data <- read.csv("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Data/SheepSampleSheet.csv")

# Setting AndroAge with predicted time of puberty onset = age minus 0.5 years 
# (i.e estimate to reach puberty at 6 months)
sheep_data$YPMP <- sheep_data$Age-0.5
sheep_data$YPMP[sheep_data$Castration_Sex=="Wether"] <- 0
sheep_data$YPMP[sheep_data$Castration_Sex=="Ewe"] <- 0

# Creating a new column for methylation in the form of a beta value
sheep_data$Beta <- sheep_data$CG_Methylation/100
```

### A birds-eye view of the data 

```{r}
# Subset based on error allowance
sheep_data_5 <- sheep_data %>% filter(CG_Calls>=497)

# Age distribution with +/- 5% error in 75% methylated samples
age_dist_5 <- ggplot(sheep_data_5, aes(x = Age)) +
  geom_histogram(aes(fill = Castration_Sex), position = "dodge", alpha = 1, 
                 bins = 20) +
  scale_fill_manual(values = get_pal("Takahe")[c(1,2,4)]) +
  theme_light() +
  labs(title = "Distribution of age by sex (Sheep ear), >497 CG calls 
       (±5% error at 75% methyated)", fill = "Sex", x="Age (Years)") +
  ylab("Frequency")
age_dist_5
```

### Plotting methylation data

```{r}
# +/- 5% error in 75% methylated samples
MKLN1_sheep_5 <- ggplot(sheep_data_5, aes(x=Age, y=CG_Methylation, 
                                          colour=Castration_Sex)) +
  geom_jitter(width=0.1, size=4, alpha=2/3) +
  geom_smooth(method="lm") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2,4)], 
                      labels = c("Female", "Intact Male", "Castrated Male")) +
  theme_light() +
  labs(colour="Sex", y="Methylation (%)", 
       x="Age (Years)")
MKLN1_sheep_5


ggsave("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Revision/FinalFigures/FigureSX_MKLN1.png",MKLN1_sheep_5, width=9, height = 6, dpi=1000)
```

# Fitting the Androgen Clock

```{r}
ram_data_5 <- sheep_data_5 %>% filter(Castration_Sex=="Ram")

# Run model, 5% error
model_5p <- lm(YPMP ~ Beta, data=ram_data_5)
summary(model_5p)
```

### Diagnostic plots

```{r}
autoplot(model_5p, which=c(1,2,3,5), size=3, alpha=4/5, colour="#313657", 
         smooth.colour="#DD3C51", smooth.linetype="dashed", label.size = 3) + 
  theme_light() + theme(plot.title = element_text(color = "white"))
```

```{r}
fitted_5 <- as.data.frame(model_5p$fitted.values)
fitted_5["YPMP"] <- ram_data_5$YPMP
fitted_5$YPMP <- as.numeric(fitted_5$YPMP)
fitted_5$Fit_AndroAge <- as.numeric(fitted_5$`model_5p$fitted.values`)
fitted_5$Age <- ram_data_5$Age
```

# Calculating error
### All

```{r}
Metrics::mdae(actual=fitted_5$YPMP, predicted=fitted_5$`model_5p$fitted.values`)
# 0.502
```

### Correlation 

```{r}
cor.test(fitted_5$YPMP, fitted_5$Fit_AndroAge, method="pearson")
# 0.912
```

# Leave-one-out cross validation

```{r}
ram_data_5$Beta <- as.numeric(ram_data_5$Beta)
ram_data_5$YPMP <- as.numeric(ram_data_5$YPMP)
ctrl <- trainControl(method="LOOCV")
model_5p_LOO <- train(YPMP ~ Beta, data=ram_data_5, method="lm", trControl=ctrl)
print(model_5p_LOO)
```

```{r}
fittedLOO_5 <- as.data.frame(model_5p_LOO$pred)
colnames(fittedLOO_5)[1] <- "Predicted_AndroAge_LOO"
colnames(fittedLOO_5)[2] <- "YPMP"

ram_data_5$Predicted_AndroAge_LOO <- fittedLOO_5$Predicted_AndroAge_LOO

Metrics::mdae(actual=ram_data_5$YPMP, predicted=ram_data_5$Predicted_AndroAge_LOO)
# 0.509

cor.test(ram_data_5$YPMP, ram_data_5$Predicted_AndroAge_LOO, method="pearson") 
# 0.909
```

```{r}
fit5_LOO <- ggplot(ram_data_5, aes(x=YPMP, y=Predicted_AndroAge_LOO, colour=Castration_Sex, fill=Castration_Sex)) +
  geom_point(size=4, alpha=3/4) + 
  geom_smooth(method="lm", se=F) +
  scale_colour_manual(values = c("#313657")) +
  geom_abline(linetype="dashed") +
  theme_light() +
  ylab("Model AndroAge (Years)") +
  xlab("Years Post Male Puberty (YPMP)") +
    theme(legend.position = "none") 
fit5_LOO
```


# Apply to whole dataset
```{r}
sheep_data_5$Predicted_AndroAge <- predict(model_5p, newdata = sheep_data_5)

Metrics::mdae(actual=sheep_data_5$Predicted_AndroAge, predicted=sheep_data_5$YPMP) # 0.362
cor.test(sheep_data_5$YPMP, sheep_data_5$Predicted_AndroAge, method="pearson") # 0.932

# calculate error
sheep_data_5$AndroAgeAccel <- sheep_data_5$Predicted_AndroAge-sheep_data_5$YPMP
```

# For each sex independently
```{r}
ram_data_5 <- sheep_data_5 %>% filter(Castration_Sex=="Ram") 
Metrics::mdae(actual=ram_data_5$Predicted_AndroAge, predicted=ram_data_5$YPMP) # 0.502
cor.test(ram_data_5$YPMP, ram_data_5$Predicted_AndroAge, method="pearson") # 0.912

wether_data_5 <- sheep_data_5 %>% filter(Castration_Sex=="Wether") 
Metrics::mdae(actual=wether_data_5$Predicted_AndroAge, predicted=wether_data_5$YPMP) # 0.281
mean(wether_data_5$Predicted_AndroAge) # 0.003

ewe_data_5 <- sheep_data_5 %>% filter(Castration_Sex=="Ewe")
Metrics::mdae(actual=ewe_data_5$Predicted_AndroAge, predicted=ewe_data_5$YPMP) # 0.360
mean(ewe_data_5$Predicted_AndroAge) # -0.226
```

# Plot

```{r}
sheep_all_plot <- ggplot(sheep_data_5, aes(x=YPMP, y=Predicted_AndroAge, colour=Castration_Sex)) +
 geom_point(size=4, alpha=3/4) + 
  geom_smooth(method="lm", se=F, colour="#000000") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2,4)]) +
  theme_light() +
  geom_abline(linetype="dashed") +
  ylab("Model AndroAge (Years)") +
  xlab("Years Post Male Puberty (YPMP)") +
  theme(legend.position = "none") 
sheep_all_plot
```

# Plot AndroAge for ewes and wethers
```{r}
EweWether_data_5 <- sheep_data_5 %>% filter(!(sheep_data_5$Castration_Sex=="Ram"))
EweWether_data_5 <- EweWether_data_5 %>% mutate(Castration_Sex = ifelse(Castration_Sex == "Ewe", "Female", Castration_Sex))
EweWether_data_5 <- EweWether_data_5 %>% mutate(Castration_Sex = ifelse(Castration_Sex == "Wether", "Castrated Male", Castration_Sex))


d2sheep <- EweWether_data_5 %>%
  dplyr::group_by(Castration_Sex) %>%
  dplyr::summarise(
    mean = mean(Predicted_AndroAge, na.rm = TRUE),
    sd = sd(Predicted_AndroAge, na.rm = TRUE)
  )

AndroAgeEweWether <- ggplot(d2sheep, aes(x=Castration_Sex, y=mean)) + 
  geom_bar(stat = "identity", fill="lightgrey", colour="black") + 
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.5) + 
  geom_jitter(EweWether_data_5, mapping = aes(x=Castration_Sex, y=Predicted_AndroAge, colour=Castration_Sex), size=4, alpha=1/2, width=0.1) +
  ylab("AndroAge (Years)") + theme_classic() + geom_hline(yintercept=0) + 
  theme_light() +
  scale_colour_manual(values = get_pal("Takahe")[c(4,1)]) +
  theme(legend.position="none", plot.title = element_text(hjust = 0.5))+
  xlab("Sex") +
  theme(text = element_text(size = 10))# +
  #ylim(-1, 2.7)
AndroAgeEweWether
ggsave("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Revision/FinalFigures/FigureSX_EweWether_AndroAge.png", AndroAgeEweWether, width=5,height=4, dpi=1000)
```



```{r}
sheep_EC_data <- read.csv("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Data/SheepEpigeneticClockData.csv")

compare_accel_data <- merge(sheep_EC_data, sheep_data_5, by.x = "ExternalSampleID", by.y = "Sample")

compare_scat <- ggplot(compare_accel_data, aes(x = AgeAccelbasedOnAll, y = AndroAgeAccel, colour = Castration_Sex)) +
  geom_point(size = 4, alpha = 3/4) +
  scale_colour_manual(values = c("#DD3C51", "#313657", "#6C90B9"), name = "Sex", labels = c("Female", "Intact male", "Castrated male")) +
  theme_light() +
  labs(colour = "Sex", y = "AndroAge acceleration (Years)", x = "DNAmAge acceleration (Years)") +
  geom_smooth(method = "lm", se = FALSE, aes(group = interaction(Castration_Sex, Sex))) +
  geom_abline(linetype = "dashed") +
  theme(legend.position = "none")
 
compare_scat
```



```{r}
legend_b <- cowplot::get_legend(
  compare_scat + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom") +
            guides(colour = guide_legend(title.position = "top", title.hjust=0.5))
)

figure2b <- cowplot::plot_grid(sheep_all_plot, compare_scat, labels=c("A", "B"), nrow=1, rel_widths = c(1,1.05))
figure2b

figure2 <- cowplot::plot_grid(figure2b, legend_b, nrow=2, rel_heights = c(1, 0.15))
figure2

ggsave("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Revision/FinalFigures/Figure2.png",figure2, width=9, height = 4.5, dpi=1000)

```

# Correlations

```{r}
# All
complete_cases_compare <- compare_accel_data[complete.cases(compare_accel_data$AgeAccelbasedOnAll, compare_accel_data$AndroAgeAccel), ]
cor.test(complete_cases_compare$AgeAccelbasedOnAll, complete_cases_compare$AndroAgeAccel, method = "pearson")
# cor=-0.114, p=0.1871 - not significant

# Ewe
complete_ewes <- complete_cases_compare %>% filter(Castration_Sex=="Ewe")
cor.test(complete_ewes$AgeAccelbasedOnAll, complete_ewes$AndroAgeAccel, method = "pearson")
# cor=-0.160, p=0.2822 - not significant 

# Ram
complete_rams <- complete_cases_compare %>% filter(Castration_Sex=="Ram")
cor.test(complete_rams$AgeAccelbasedOnAll, complete_rams$AndroAgeAccel, method = "pearson")
# cor=-0.119, p=0.4591 - not significant 

# Wether
complete_wethers <- complete_cases_compare %>% filter(Castration_Sex=="Wether")
cor.test(complete_wethers$AgeAccelbasedOnAll, complete_wethers$AndroAgeAccel, method = "pearson")
# cor=0.123, p=0.4096 - not significant 

##### no significant correlations in any case. #####
```


does androage have anything to do with dnam age acceleration?

```{r}
compare_scat2 <- ggplot(compare_accel_data, aes(x = AgeAccelbasedOnAll, y = Predicted_AndroAge, colour = Castration_Sex)) +
  geom_point(size = 4, alpha = 3/4) +
  scale_colour_manual(values = c("#DD3C51", "#313657", "#6C90B9"), name = "Sex", labels = c("Female", "Intact male", "Castrated male")) +
  theme_light() +
  labs(colour = "Sex", y = "AndroAge (Years)", x = "DNAmAge Acceleration (Years)") +
  geom_smooth(method = "lm", se = FALSE, aes(group = interaction(Castration_Sex, Sex))) +
  geom_abline(linetype = "dashed") +
  theme(legend.position = "bottom")
 
compare_scat2

ggsave("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Revision/FinalFigures/FigureSX_AndroAgevsDNAmAgeAccel.png",compare_scat2, width=6, height = 5, dpi=1000)
```

# Correlations

```{r}
# All
complete_cases_compare <- compare_accel_data[complete.cases(compare_accel_data$AgeAccelbasedOnAll, compare_accel_data$Predicted_AndroAge), ]
cor.test(complete_cases_compare$AgeAccelbasedOnAll, complete_cases_compare$Predicted_AndroAge, method = "pearson")
# cor=0.092, p=0.2862 - not significant

# Ewe
complete_ewes <- complete_cases_compare %>% filter(Castration_Sex=="Ewe")
cor.test(complete_ewes$AgeAccelbasedOnAll, complete_ewes$Predicted_AndroAge, method = "pearson")
# cor=-0.160, p=0.2822 - not significant 

# Ram
complete_rams <- complete_cases_compare %>% filter(Castration_Sex=="Ram")
cor.test(complete_rams$AgeAccelbasedOnAll, complete_rams$Predicted_AndroAge, method = "pearson")
# cor=0.4178, p=0.006565 - significant ! 

# Wether
complete_wethers <- complete_cases_compare %>% filter(Castration_Sex=="Wether")
cor.test(complete_wethers$AgeAccelbasedOnAll, complete_wethers$Predicted_AndroAge, method = "pearson")
# cor=0.123, p=0.4096 - not significant 
```

