---
title: "Sheep Androgen Clock Construction - Array"
author: "Victoria Sugrue"
date: 'Last updated: 10 July 2023'
output:
  word_document:
    toc: yes
    toc_depth: '6'
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 6
  pdf_document:
    toc: yes
    toc_depth: 5
    df_print: kable
    latex_engine: xelatex
subtitle: ' '
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load required packages
```{r}
library(ggplot2)
library(caret)
library(dplyr)
library(Manu)
library(knitr)
library(lmtest)
library(ggfortify)
```

# Data preparation
```{r}
#read in minfi normalised beta values
sheep_array_cg215241162 <- read.csv("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Data/array_cg21524116.csv", header=FALSE)

#transpose the dataframe and remove unneeded columns
sheep_array_cg21524116 <- as.data.frame(t(sheep_array_cg215241162[,-c(1:2)]))

#read in file for headers and transpose it, assign the headers to the minfi normalised data
headers <- read.csv("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/Data/array_headers.csv", header=FALSE)
theaders <- as.list(t(headers))
names(sheep_array_cg21524116) <- theaders

#remove blood
sheep_array_cg21524116_ear <- sheep_array_cg21524116 %>% filter(!(Sex=="Blood"))
#remove "CanBeUsedInAgingStudied==NO"
sheep_array_cg21524116_ear <- sheep_array_cg21524116_ear %>% filter(!(ExternalSampleID=="RAM_H8_2764_2018"))
sheep_array_cg21524116_ear <- sheep_array_cg21524116_ear %>% filter(!(ExternalSampleID=="RAM_H5_2777_2018"))

# Setting AndroAge with predicted time of puberty onset = age minus 0.5 years (i.e estimate to reach puberty at 6 months)
sheep_array_cg21524116_ear$YPMP <- as.numeric(sheep_array_cg21524116_ear$Age)-0.5
sheep_array_cg21524116_ear$YPMP[sheep_array_cg21524116_ear$Sex=="Wether"] <- 0
sheep_array_cg21524116_ear$YPMP[sheep_array_cg21524116_ear$Sex=="Ewe"] <- 0

# Subset by sex/castration status
ram_array_ear_data <- sheep_array_cg21524116_ear %>% filter(Sex=="Ram")
wether_array_ear_data <- sheep_array_cg21524116_ear %>% filter(Sex=="Wether")
ewe_array_ear_data <- sheep_array_cg21524116_ear %>% filter(Sex=="Ewe")

```

# A birds-eye view of the data 

```{r}
# Age distribution
age_dist_ARRAY <- ggplot(sheep_array_cg21524116_ear, aes(x = as.numeric(Age))) +
  geom_histogram(aes(fill = Sex), position = "dodge", alpha = 1, bins = 20) +
  scale_fill_manual(values = get_pal("Takahe")[c(1,2,4)]) +
  theme_light() +
  labs(title = "Distribution of age by sex (Sheep ear, Array)", fill = "Sex", x="Age (Years)") +
  ylab("Frequency")
age_dist_ARRAY
```

# Plotting methylation data

```{r}
MKLN1_sheep_ARRAY <- ggplot(sheep_array_cg21524116_ear, aes(x=as.numeric(Age), y=as.numeric(cg21524116)*100, colour=Sex)) +
  geom_jitter(width=0.1, size=4, alpha=2/3) +
  geom_smooth(method="lm") +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2,4)]) +
  theme_light() +
  labs(title = "Methylation at cg21524116, Array", colour="Sex", y="Methylation (%)", x="Age (Years)")
MKLN1_sheep_ARRAY
```


# Fitting model to array data, sheep ear only

```{r}
# Run model, array data, training on rams
model_array1 <- lm(YPMP ~ as.numeric(cg21524116), data=ram_array_ear_data)
summary(model_array1)

```

# Diagnostic plots

```{r}
autoplot(model_array1, which=c(1,2,3,5), size=3, alpha=4/5, colour="#313657", smooth.colour="#DD3C51", smooth.linetype="dashed", label.size = 3) + theme_light() + theme(plot.title = element_text(color = "white"))
```




```{r}
fitted_array <- as.data.frame(model_array1$fitted.values)
fitted_array["YPMP"] <- ram_array_ear_data$YPMP
fitted_array$YPMP <- as.numeric(fitted_array$YPMP)
fitted_array$Fit_AndroAge <- as.numeric(fitted_array$`model_array1$fitted.values`)
fitted_array$Age <- ram_array_ear_data$Age
```

# Calculating error
## All

```{r}
MAE_model_array1 <- as.data.frame(Metrics::mdae(actual=fitted_array$YPMP, predicted=fitted_array$Fit_AndroAge))
MAE_model_array1 # 0.228
```


# Testing the model fit further: 

# Breusch-Pagan test of variance (homo/heteroscedasticity)

```{r}
plot(model_array1$residuals)
bptest(model_array1)
```

# 0.4728

# Shapiro-Wilk test of normality

```{r}
shapiro.test(model_array1$residuals)
```

# 9.55e-05

# Leave-one-out cross validation

```{r}
ram_array_ear_data$cg21524116 <- as.numeric(ram_array_ear_data$cg21524116)
ram_array_ear_data$YPMP <- as.numeric(ram_array_ear_data$YPMP)
ctrl <- trainControl(method="LOOCV")
RamLOO_ARRAY <- train(YPMP ~ cg21524116, data=ram_array_ear_data, method="lm", trControl=ctrl)
summary(RamLOO_ARRAY)
```

```{r}
print(RamLOO_ARRAY)
```

```{r}
fittedLOO_ARRAY <- as.data.frame(RamLOO_ARRAY$pred)
fittedLOO_ARRAY["YPMP"] <- ram_array_ear_data$YPMP

RamLOO_ARRAY_MAE <- as.data.frame(Metrics::mdae(actual=fittedLOO_ARRAY$YPMP, predicted=fittedLOO_ARRAY$pred))
RamLOO_ARRAY_MAE # 0.231
```

```{r}
cor(fittedLOO_ARRAY$obs, fittedLOO_ARRAY$pred, method="pearson")
# 0.941


ramLOOfit_array <- ggplot(fittedLOO_ARRAY, aes(x=obs, y=pred)) +
  geom_point(size=4, alpha=3/4, colour="#313657") + 
  geom_smooth(method="lm", se=F, colour="#DD3C51") +
  geom_abline(linetype="dashed") +
  theme_light() +
  ylab("Model AndroAge (Years)") +
  xlab("Years Post Male Puberty (YPMP)") +
  labs(title = "Ram androgen clock fit (Array)",
              subtitle = "MAE=0.231, cor=0.941") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5))
ramLOOfit_array
```




# Now, to include ewes and wethers.

```{r}
ewe_array_ear_data$Model_AndroAge <- predict(model_array1, newdata = ewe_array_ear_data)

ewe_MAE_ARRAY <- as.data.frame(Metrics::mdae(actual=ewe_array_ear_data$YPMP, predicted=ewe_array_ear_data$Model_AndroAge))
ewe_MAE_ARRAY
# Ewe MAE = 0.255 years
```


```{r}
wether_array_ear_data$Model_AndroAge <- predict(model_array1, newdata = wether_array_ear_data)

wether_MAE_ARRAY <- as.data.frame(Metrics::mdae(actual=wether_array_ear_data$YPMP, predicted=wether_array_ear_data$Model_AndroAge))
wether_MAE_ARRAY
# Wether MAE = 0.314 years

ewewether_array <- rbind(ewe_array_ear_data, wether_array_ear_data)

ewewether_array_violin <- ggplot(ewewether_array, aes(x=Sex, y=Model_AndroAge, fill=Sex)) +
  geom_violin() + 
  geom_point(size=4, alpha=3/4, shape=1) +
  theme_light() +
    scale_fill_manual(values = get_pal("Takahe")[c(1,4)]) +
  geom_hline(yintercept = 0, linetype="dashed") +
  ylab("AndroAge predicted by model (Years)") +
  xlab("Sex") +
  ylab("Model AndroAge (Years)") +
  xlab("Sex") +
  labs(title = "Ewe and wether androgen clock fit (Array), trained on all rams",
              subtitle = "Ewe MAE=0.255, Wether MAE=0.314") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5)) +
  theme(legend.position = "none")
ewewether_array_violin

```

# LOOCV across whole dataset
```{r}
sheep_array_cg21524116_ear$cg21524116 <- as.numeric(sheep_array_cg21524116_ear$cg21524116)
sheep_array_cg21524116_ear$YPMP <- as.numeric(sheep_array_cg21524116_ear$YPMP)
ram_array_ear_data$cg21524116 <- as.numeric(ram_array_ear_data$cg21524116)
ram_array_ear_data$YPMP <- as.numeric(ram_array_ear_data$YPMP)

ctrl <- trainControl(method="LOOCV")
model_array_LOO_all <- train(YPMP ~ cg21524116, data = ram_array_ear_data, method = "lm", trControl = ctrl)

print(model_array_LOO_all)

predicted_YPMP <- predict(model_array_LOO_all, newdata = sheep_array_cg21524116_ear)
median(abs(predicted_YPMP - sheep_array_cg21524116_ear$YPMP))
# 0.263

cor(sheep_array_cg21524116_ear$YPMP, sheep_array_cg21524116_ear$Model_AndroAge, method="pearson")
# 0.890
```

```{r}
sheep_array_cg21524116_ear$Model_AndroAge <- predict(model_array1, newdata = sheep_array_cg21524116_ear)

sheeparray_paper <- ggplot(sheep_array_cg21524116_ear, aes(x=YPMP, y=Model_AndroAge, colour=Sex, fill=Sex)) +
  geom_point(size=4, alpha=3/4) + 
  geom_smooth(method="lm", se=F) +
  scale_colour_manual(values = c("#DD3C51", "#313657", "#6C90B9"), name= "Sex", labels = c("Female", "Castrated male", "Intact male")) +
  geom_abline(linetype="dashed") +
  theme_light() +
  ylab("Model AndroAge (Years)") +
  xlab("Years Post Male Puberty (YPMP)") +
  theme(legend.position = "bottom") +
  labs(title = "Array-based androgen clock fit, sheep",
              subtitle = "MAE=0.263, cor=0.890") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5)) +
    guides(fill = FALSE)

sheeparray_paper

ggsave("/Users/victoriasugrue/Dropbox (Otago University)/Androgen_Clock/Manuscript/sheep_array_fit.png",sheeparray_paper, width=6, height = 6, dpi=1000)

```




# Plot AndroAge acceleration

```{r}
ram_array_ear_data$Model_AndroAge <- predict(model_array1, newdata = ram_array_ear_data)

combined_array <- rbind(ram_array_ear_data, wether_array_ear_data, ewe_array_ear_data)

combined_array$AndroAgeAccel <- combined_array$Model_AndroAge-combined_array$YPMP

write.csv(combined_array, "/Users/victoriasugrue/Dropbox (Otago University)/combined_array.csv")
d2combined_array <- read.csv("/Users/victoriasugrue/Dropbox (Otago University)/d2_combined_array.csv")

SheepAndroAgeAccel_Array <- ggplot(d2combined_array, aes(x=Group, y=mean)) + 
  geom_bar(stat = "identity", fill="lightgrey", colour="black") + 
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.5) + 
  geom_jitter(combined_array, mapping = aes(x=Sex, y=AndroAgeAccel, colour=Sex), size=4, alpha=1/2, width=0.05) +
  ylab("AndroAge Acceleration (Years)") + theme_classic() + geom_hline(yintercept=0) + 
  theme_light() +
  scale_colour_manual(values = get_pal("Takahe")[c(1,2,4)]) +
  theme(legend.position="none") 
SheepAndroAgeAccel_Array
```
